
% !TEX root = scombinatorics.tex
\documentclass[scombinatorics.tex]{subfiles}
\begin{document}
\chapter*{References}
\addcontentsline{toc}{chapter}{References}

\begin{biblist}[]\normalsize

  \bib{AK1}{article}{
    author={Alon, N.},
    author={Kleitman, D. J.},
    title={A purely combinatorial proof of the Hadwiger Debrunner $(p,q)$
    conjecture},
    % note={The Wilf Festschrift (Philadelphia, PA, 1996)},
    journal={Electron. J. Combin.},
    volume={4},
    date={1997},
    % number={2},
    % pages={Research Paper 1, approx. 8},
    % review={\MR{1444148}},
 }

 \bib{AK2}{article}{
    author={Alon, Noga},
    author={Kleitman, Daniel J.},
    title={Piercing convex sets and the Hadwiger-Debrunner $(p,q)$-problem},
    journal={Adv. Math.},
    volume={96},
    date={1992},
    % number={1},
    % pages={103--112},
    % issn={0001-8708},
    % review={\MR{1185788}},
    % doi={10.1016/0001-8708(92)90052-M},
 }
     
\bib{ARS}{article}{
   author={Anstee, R. P.},
   author={R\'onyai, Lajos},
   author={Sali, Attila},
   title={Shattering news},
   journal={Graphs Combin.},
   volume={18},
   date={2002},
%  number={1},
%  pages={59--73},
%  issn={0911-0119},
%  doi={10.1007/s003730200003},
}
\bib{DGL}{book}{
   author={Devroye, Luc},
   author={Gy\"{o}rfi, L\'{a}szl\'{o}},
   author={Lugosi, G\'{a}bor},
   title={A probabilistic theory of pattern recognition},
   publisher={Springer-Verlag},
   date={1996},
}
\bib{DL}{book}{
   author={Devroye, Luc},
   author={Lugosi, G\'{a}bor},
   title={Combinatorial methods in density estimation},
   series={Springer Series in Statistics},
   publisher={Springer-Verlag},
   date={2001},
}
\bib{gowers}{article}{
   author={Gowers, Timothy},
   title={\href{https://gowers.wordpress.com/2008/07/31/dimension-arguments-in-combinatorics}{Dimension arguments in combinatorics}},
   journal={Gowers's Weblog},
   date={2008}
}
\bib{kalai}{article}{
  author={Kalai, Gil} ,
  title={\href{https://gilkalai.wordpress.com/2008/09/28/extremal-combinatorics-iii-some-basic-theorems}{Extremal Combinatorics III: Some Basic Theorems}},
  journal={Combinatorics and more},
  date={2008}
}
\bib{matousek}{article}{
   author={Matou\v{s}ek, Ji\v{r}\'{\i}},
   title={Bounded VC-dimension implies a fractional Helly theorem},
   journal={Discrete Comput. Geom.},
   volume={31},
   date={2004},
  %  number={2},
  %  pages={251--255},
  %  issn={0179-5376},
  %  review={\MR{2060639}},
  %  doi={10.1007/s00454-003-2859-z},
}
\bib{LPmatousek}{book}{
  author={Matousek, J\v{\i}ri},
  author={Gartner, Bernd},
  title={Understanding and Using Linear Programming},
  publisher={Springer-Verlag},
  date={2007},
  pages={vi+222},
}
\bib{pajor}{book}{
   author={Pajor, Alain},
   title={Sous-espaces $l^n_1$ des espaces de Banach},
%   language={French},
   series={Travaux en Cours [Works in Progress]},
   volume={16},
%   note={With an introduction by Gilles Pisier},
   publisher={Hermann, Paris},
   date={1985},
%   pages={xii+112},
%   isbn={2-7056-6021-6}
}
\bib{sauer}{article}{
   author={Sauer, N.},
   title={On the density of families of sets},
   journal={J. Combinatorial Theory Ser. A},
   volume={13},
   date={1972},
   pages={145--147},
}
\bib{shelah72}{article}{
   author={Shelah, Saharon},
   title={A combinatorial problem; stability and order for models and theories in infinitary languages},
   journal={Pacific J. Math.},
   volume={41},
   date={1972},
   pages={247--261},
}
\bib{VC}{article}{
   author={Vapnik, V. N.},
   author={Chervonenkis, A. Ya.},
   title={On the uniform convergence of relative frequencies of events to their probabilities},
   note={Reprint of Theor. Probability Appl. {\bf 16} (1971), 264--280},
   conference={title={Measures of complexity},},
   book={publisher={Springer, Cham},},
   date={2015},
   pages={11--30},
}
\end{biblist}
\end{document}