% !TEX root = scombinatorics.tex
\documentclass[scombinatorics.tex]{subfiles}
\begin{document}
\chapter{Stability}
\label{sauer}



\def\medrel#1{\parbox[t]{5ex}{$\displaystyle\hfil #1$}}
\def\ceq#1#2#3{\parbox[t]{20ex}{$\displaystyle #1$}\medrel{#2}{$\displaystyle #3$}}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The order property}\label{ladder}

The \emph{chain index\/} of $\phi(\U\,;b)_{b\in\V}$, or of $\phi(x\,;z)$ when $\U$ and $\V$ are clear, is the maximal length (that is, $n+1$) of a chain of the form

\ceq{\ssf{ch}\hfill\phi(A\,;b_0)\ \subset}{\dots}{\subset\ \phi(A\,;b_n)}

for some set $A\subseteq\U$ and some $b_0,\dots,b_n\in\V$.
Note that we allow $\phi(A\,;b_0)$ to be empty.
This choice produces a small asymmetry in the definition of ladder; see also Fact~\ref{fact_stability_dual}.

\begin{example}
  If $\phi(\U\,;b)_{b\in\V}$ consists of just one set, the ladder index is $1$.
  If it contains two distinct sets, the  chain index is at least $2$ and it is exactly $2$ if there are no more two sets, or if all sets are disjoint.\QED
\end{example}

If a maximal length does not exist, we say that $\phi(x\,;z)$ is \emph{unstable,} or that it has the \emph{order-property.} 
Otherwise we say that it is \emph{stable.}

In place of requiring the existence of the chain in \ssf{ch}, we could equivalently ask for a pair of tuples $a_1,\dots,a_n\in\U$ and $b_0,\dots,b_n\in\V$ such that

\ceq{\ssf{ld}\hfill\phi(a_h\,;b_k)}
{\IFF}
{h\le k.}

We call this pair of tuples a \emph{ladder\/} of length $n+1$.
We may also say \emph{ladder index\/} instead of chain index.
Setting $A=\{a_1,\dots,a_n\}$ we easily obtain a chain from a ladder, the converse is left as an easy exercise for the reader.

\begin{exercise}
  Let $\phi(x\,;z)$ have chain index $n+1$ or more. 
  Let $A\subseteq\U$ be a minimal set such that a chain as in \ssf{ch} obtains for some $b_0,\dots,b_n\in\V$.
  Prove that there is a ladder $a_1,\dots,a_n$ and $b_0,\dots,b_n$ such that $A=\{a_1,\dots,a_n\}$ and conclude that $\phi(x\,;z)$ has ladder index $n+1$.\QED
\end{exercise}

The following facts are obvious but worth noting.

\begin{fact}\label{fact_stability_dual}
  Let $\phi(x\,;z)$ have ladder index $n+1$. 
  Then $\phi(x\,;z)^{\rm op}$ has ladder index $\ge n$.\QED
\end{fact}
 
\begin{fact}\label{fact_stability_neg}
  Let $\phi(x\,;z)$ have ladder index $n$. 
  Then $\neg\phi(x\,;z)$ has ladder index $n$.\QED
\end{fact}

\begin{proof}
  If for all $h\in(n]$ and $k\in[n]$ 

  \ceq{\hfill\neg\phi(a_h\,;b_k)}
  {\IFF}
  {h\le k}

  then $a'_h=a_{n+1-h}$ and $b'_k=b_{n-k}$ satisfy

  \ceq{\hfill\phi(a'_h\,;b'_k)}
  {\IFF}
  {\phi(a_{n+1-h}\,;b_{n-h})}

  \ceq{}
  {\IFF}
  {n-k>n+1-h}

  \ceq{}
  {\IFF}
  {h\le k}
\end{proof}
 
The following definition is connected with those above, though in a less evident manner.
We write ${}^n2$ for the set of binary sequences of length $n$ or, more precisely, the set of functions $s:[n)\to[2)$.
We write $s_h$ for the value of $s$ at $h$, and $s{\restriction} h$ for the restriction of $s$ to $[h)$.
We define ${}^{<n}2=\big\{r\, :\, r\in {}^h2,\ h\in[n)\big\}$.

A \emph{branching tree\/} of hight $n$ for the formula $\phi(x\,;z)$ is a function 

\ceq{\hfill\bar{a}\ :\ {}^{<n}2}{\to}{\U}
\nopagebreak[4]\par
\ceq{\hfill r}{\mapsto}{a_r},

which we may also present by writing $\bar a=\<a_r:r\in{}^{<n}2\>$, such that

\ceq{\ssf{2r}\hfill\0}
{\neq}
{\bigcap_{h=0}^{n-1}\neg^{s_h}\,\phi(a_{s\restriction h}\,;\V)}\hfill for all $s\in {}^n2$.

where $\neg^i$, for $i$ a non negative integer, denotes a negation symbol repeated $i$ times.
In other words \ssf{2r} requires the existence of some $\<b_s:s\in{}^n2\>$ such that

\ceq{\hfill\phi(a_{s\restriction h}\,;b_s)}
{\IFF}
{s_h=1}\hfill for all pairs $s\in {}^n2$ and $h\in[n)$

or, with slightly different notation,

\ceq{\hfill\phi(a_r\,;b_s)}
{\IFF}
{r^\frown1\subseteq s}\hfill for all pairs $r\subset s\in {}^n2$.

It helps to represent a branching tree as follows.
For definiteness, fix $n=3$.
Consider a full binary tree of height $n+1$ and assign to each internal node (different from root and leaves) a formula as depicted below.
Then \ssf{2r} requires that all formulas in each branch $s\in2^n$ are satified by some $b_s\in\V$.
\medskip

% Set the overall layout of the tree
\tikzstyle{level 1}=[level distance=3.5cm, sibling distance=2.5cm]
\tikzstyle{level 2}=[level distance=3.5cm, sibling distance=1.2cm]
\tikzstyle{level 3}=[level distance=2.5cm, sibling distance=0.5cm]
\tikzstyle{level 4}=[level distance=0.5cm, sibling distance=0.5cm]

% Define styles for bags and leafs
\tikzstyle{bag0} = [text width=2.5ex, align=left]
\tikzstyle{bag} = [text width=7.5ex, align=right]
%\tikzstyle{end} = [circle, minimum width=3pt,fill, inner sep=0pt]

\begin{tikzpicture}[grow=right]
\node[bag0] {{$\top$}}
    child {
        node[bag] {$\phi(a_{\0};z)$}      
            child {
                node[bag] {$\phi(a_1;z)$}
                    child {
                       node[bag] {\footnotesize$\phi(a_{11};z)$\rlap{ ----- $b_{111}$}}
                       edge from parent
                    }    
                    child {
                       node[bag] {\footnotesize$\llap{$\neg$}\phi(a_{11};z)$\rlap{ ----- $b_{110}$}}
                       edge from parent
                    }  
                 edge from parent
            }
            child {
                node[bag] {\llap{$\neg$}$\phi(a_1;z)$}
                edge from parent
                    child {
                       node[bag] {\footnotesize$\phi(a_{10};z)$\rlap{ ----- $b_{101}$}}
                       edge from parent
                    }    
                    child {
                       node[bag] {\footnotesize\llap{$\neg$}$\phi(a_{10};z)$\rlap{ ----- $b_{100}$}}
                       edge from parent
                    }  
                 edge from parent
            }
       edge from parent 
    }
    child {
        node[bag] {\llap{$\neg$}$\phi(a_{\0};z)$}         
            child {
                node[bag] {$\phi(a_0;z)$}
                    child {
                       node[bag] {\footnotesize$\phi(a_{01};z)$\rlap{ ----- $b_{011}$}}
                       edge from parent
                    }    
                    child {
                       node[bag] {\footnotesize\llap{$\neg$}$\phi(a_{01};z)$\rlap{ ----- $b_{010}$}}
                       edge from parent
                    }  
                 edge from parent
            }
            child {
                node[bag] {\llap{$\neg$}$\phi(a_0;z)$}
                edge from parent
                    child {
                       node[bag] {\footnotesize$\phi(a_{00};z)$\rlap{ ----- $b_{001}$}}
                       edge from parent
                    }    
                    child {
                       node[bag] {\footnotesize\llap{$\neg$}$\phi(a_{00};z)$\rlap{ ----- $b_{000}$}}
                       edge from parent
                    }  
                 edge from parent
            } 
        edge from parent
    };
\end{tikzpicture}
\smallskip

The Shelah local \emph{2-rank\/} of $\phi(x\,;z)$ is the maximal height of a branching tree for $\phi(x\,;z)$.
If such a maximal integer does not exist, we say that the 2-rank is infinite.

A branching tree $\bar a'=\<a'_r:r\in{}^m2\>$ is a \emph{subtree\/} of $\bar a$ if there is an $\subseteq$-preserving map $f:{}^m2\to{}^n2$ such that $a'_r=a_{fr}$.

For the theorem below we need the following Ramsey-like lemma.

% \begin{lemma}\label{lem_Ramsey_on_trees}
%   Let $\bar a=\<a_r\, :\, r\in{}^{<2m}2\>$ be a branching tree for $\phi(x\,;z)$.
%   Let $k\in[2m]$ be given.
%   Then, for every $2$-coloring of $\range(\bar a)$, there is a branching tree $\bar a'=\<a'_r\, :\, r\in{}^m2\>$ such that  $\range(\bar a')$ is a monochromatic subset of $\range(\bar a)$.\QED
% \end{lemma}

% As it happens, a more general lemma is easier to prove.

\begin{lemma}\label{lem_Reamsey_tree}
  Let $\bar a=\<a_r\, :\, r\in{}^{<n}2\>$ be a branching tree for $\phi(x\,;z)$.
  Let $k\in[n]$ be given.
  Then, for every red-blue coloring of $\range(\bar a)$,  there is a monochromatic branching subtree, say $\bar a'=\<a'_r\, :\, r\in{}^{n'}2\>$, and either all nodes of $\bar a'$ are red and $n'=k$, or they are all blue and $n'=n-k$.
\end{lemma}

\begin{proof}
  Induction on $n$.
  First, note that the lemma is trivial when $k=0$ or $k=n$.
  In particular the lemma holds for $n=0$.

  Assume the lemma true for $n$ and prove it for $n+1$.
  Let  $\bar a=\<a_r\, :\, r\in{}^{n+1}2\>$ be given.
  Fix some $k\in[n+1]$. 
  We want a branching tree $\bar a'$ that is either red of height $k$ or blue of height $n+1-k$.
  
  If we discard the trivial cases, we can assume $k\in(n]$.

  For $i=0,1$ we define $\bar a_i=\<a_{i^\frown r}\;:\;r\in {}^{<n}2\>$.

  First suppose that $a_\0$ is blue.
  If, for either $i=0$ or $i=1$, there is a red branching tree $\bar a'_i$ of height $k$, we are done.
  Otherwise, for both $i=0,1$ there is a blue branching tree $\bar a'_i$ of hight $n-k$.
  Then we graft these two trees on $a_\0$, which is also blue, and obtain the required blue tree of height $n-k+1$.
  
  Now suppose that $a_\0$ is red.
  If, for either $i=0$ or $i=1$, there is a blue branching tree $\bar a'_i\subseteq \bar a_i$ of height $n-(k-1)$, we are done.
  Otherwise, we graft on $a_\0$ two red trees of hight $k-1$ to obtain a red tree of height $k$.  
\end{proof}

We are now ready to characterize stability via the 2-rank.
The proof is based on Hodges~\cite{hodges}.
It is a direct proof which yields an explicit bound on the 2-rank given the ladder index (it yields also the converse, but this is easy).
However, this bound seems to be far from optimal.
Shelah's original proof is model theoretic and does not gives explicit bounds.
In the next section we will introduce some of the ideas used in Shelah's proof.
% In the next section we sketch a second proof which elaborates on Shelah's original ideas.
% This proof is indirect but.

\begin{theorem}\label{thm_hodges}
  The following are equivalent
  \begin{itemize}
    \item[1.] $\phi(x\,;z)$ is stable;
    \item[2.] $\phi(x\,;z)$ has finite 2-rank.
  \end{itemize}
  Precisely, if $n_{\rm ld}$ and $n_{\rm 2r}$ are the ladder index and the 2-rank, respectively, then 
  
  \ceq{\hfill n_{\rm ld}}{<}{2^{n_{\rm 2r}+1};}

  \ceq{\hfill n_{\rm 2r}}{<}{2^{n_{\rm ld}+1}-2.}
\end{theorem}

\begin{proof}[Proof (\kern.1ex\ssf{2}\kern.3ex\boldmath$\IMP$\ssf{1})]
  % \def\ceq#1#2#3{\parbox[t]{29ex}{$\displaystyle #1$}\medrel{#2}{$\displaystyle #3$}}
  % \ssf{1}$\IMP$\ssf{2} 
  % We prove the contrapositive.
  % If there is a binary tree of hight $n$ then there is a ladder of length $n$.
  % This also proves the first inequality of the proposition.

  % Let $\{a_s\; :\; s\in2^{<n}\}$ and $\{b_s\; :\; s\in2^{n}\}$ satisfy \ssf{2r}.
  % Apply \ssf{2r} to sequences of the form $1^k\,0^{n-k}$ for $0\le k\le n$.

  % % Define $a'_1,\dots,a'_n$ and $b'_0,\dots,b'_n$, where $a'_h=a_{1^{h-1}}$ and $b'_k=b_{1^k0^{n-k}}$.
  % % By \ssf{2r} we have

  % \ceq{\hfill\phi(a_{1^k\,0^{n-k}\,\restriction\, h-1}\;;\,b_{1^k\,0^{n-k}})}
  % {\IFF}
  % {\big(1^k0^{n-k}\big)_{h-1}=1}

  % \ceq{}
  % {\IFF}
  % {h\le k}

  
  We prove the contrapositive.
  We show that if there is a ladder of length $m=2^n$, say $a_1,\dots,a_{m-1}$ and $b_0,\dots,b_{m-1}$, then there is a branching tree $\bar a'$ of height $n$.
  This also proves the first inequality above.
  In fact, if $n_{\rm ld}\ge 2^{n_{\rm 2r}+1}$, there would exist a branching tree of height $n_{\rm 2r}+1$ which is a contradiction.
  
  The branching tree $\bar a'=\<a'_r\, :\, r\in{}^{<n}2\>$ is defined as follows

  \quad $a'_r=a_h$\quad  where $h$ is obtained reading $r^\frown1^\frown0^{n-|r|-1}$ as an $n$-digit binary number.

  To verify \ssf{2r} we define for $s\in{}^n2$ 
  
  \quad $b'_s=b_k$\quad  where $k$ is obtained reading $s$ as an $n$-digit binary number.

  Then it is easy to verify that for all pairs $r\subset s\in{}^n2$

  \ceq{\hfill\phi(a'_r\,;b'_s)}
  {\IFF}
  {\phi(a_h\,;b_k)}\hfill where $h$ and $k$ are like above

  \ceq{}
  {\IFF}
  {h\le k}

  \ceq{}
  {\IFF}
  {r^\frown1^\frown0^{n-|r|-1}\ \le\ s}\hfill  as $n$-digit binary numbers

  \ceq{}
  {\IFF}
  {r^\frown1\ \subseteq\ s}

  \textbf{(\ssf{1}\kern.2ex\boldmath$\IMP$\ssf{2}\kern.1ex)}\ 
   We prove the contrapositive.
  We claim that if there is a branching tree $\bar a$ of height $2^n-2$ then there is a ladder $a_1,\dots,a_{n-1}$ and $b_0,\dots,b_{n-1}$, of length $n$, s.t. $\{b_0,\dots,b_{n-1}\}\subseteq \<b_s:s\in{}^{2^n-2}2\>$.
  This yields also the second inequality of the theorem.
  In fact, if $n_{\rm 2r}\ge 2^{n_{\rm ld}+1}-2$, there would exist a ladder of length $n_{\rm ld}+1$ which is a contradiction.

  The claim is true for $n=2$.
  Now we assume the claim is true for $n$ and prove it for $n+1$.

  Let $\<a_r\, :\, r\in{}^{<\, 2m+2}2\>$, where $m=2^n-2$, be a branching tree of height $2^{n+1}-2$.
  To each $b\in\<b_s:s\in{}^{2m+2}2\>$ we associate a red-blue coloring of $\range(\bar a)$ as follows.
  A node $a\in\range(\bar a)$ is colored
  
  \qquad\qquad \llap{red}\qquad if $\phi(a\,;b)$ holds;
  
  \qquad\qquad \llap{blue}\qquad otherwise, that is, $\neg\phi(a\,;b)$.
  
  We consider two cases that are exhaustive by Lemma~\ref{lem_Reamsey_tree}.
  Note that we are applying the lemma only to the subtree $\<a_{1^\frown r}:r\in{}^{2m+1}2\>$.

  Case 1: for some $b$ there is a red subtree of $\<a_{1^\frown r}:r\in{}^{2m+1}2\>$ of height $m+1$.
  Let $\bar a'$ be this red tree and consider its subtree $\<a'_{0^\frown r}\,:\,r\in {}^{<\,m}2\>$.
  By induction hypothesis, there are $A\subseteq\{a'_{0^\frown r}\,:\,r\in {}^{<\,m}2\}$ and $b_0,\dots,b_n$ such that

  \ceq{(1)\hfill\phi(A\,;b_0)\ \subset}{\dots}{\subset\ \phi(A\,;b_n)}

  Let $A'=A\cup\{a'_\0\}$ then

  \ceq{(2)\hfill\phi(A'\,;b_0)\ \subset}{\dots}{\subset\ \phi(A'\,;b_n)}

  In fact, as $b_0,\dots,b_n\in\neg\phi(a'_\0\,;\V)$, this is the same chain as (1).
  Therefore, if we extend the chain on the right with $\phi(A'\,;b)=A'$, we obtain the required chain of length $n+2$.
  
  Case 2: for every $b$ there is a blue subtree of $\<a_{1^\frown r}:r\in{}^{2m+1}2\>$ of height $m$.
  Choose $b\in\<b_s:s\in{}^{2m+2}2\>$ such that $\neg\phi(a_\0, b)$ and let $\bar a'$ the corresponding blue subtree.
  Apply the induction hypothesis to obtain $A\subseteq\range(\bar a')$ and $b_0,\dots,b_n$ such that (1).
  We claim that (2) above holds with $A'=A\cup\{a_\0\}$.
  In fact, $b_0,\dots,b_n\in\phi(a_\0\,;\V)$ so (2) is the chain in (1) with all sets augmented by $a_\0$.
  We can extend the chain on the left with $\phi(A'\,;b)=\0$ and obtain the required chain of length $n+2$.
\end{proof}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Approximable sets}

The notion of approximable set is a combinatorial counterpart of the model theoretical notion of externally definable set.

\begin{definition}\label{def_approx}
  We say that $\Aa\subseteq\U$ is \emph{approximable\/} if for every finite set $A\subseteq\U$ there is a $b\in\V$ such that $\phi(A\,;b)=\Aa\cap A$.
  If we also have that $\phi(\U\,;b)\subseteq\U$, then we say that $\Aa$ is approximable \emph{from below.}
\end{definition}

The following is immediate.

\begin{fact}
  The following are equivalent for every $\Aa\subseteq\U$
  \begin{itemize}
    \item[1.] $\Aa$ is approximable from below;
    \item[2.] for every finite set $A\subseteq\Aa$ there is a $b\in\V$ such that $A\subseteq\phi(\U\,;b)\subseteq\Aa$.\QED
  \end{itemize}
\end{fact}

\begin{proposition}
  Let $\phi(x\,;z)$ be a stable formula.
  Every set $\Aa\subseteq\U$ approximable by $\phi(x\,;z)$ is approximable from below by the formula

  \ceq{\hfill \psi(x\,;z_0,\dots,z_n)}{=}{\bigwedge_{i=0}^n\phi(x\,;z_i)}
  
  where $n+1$ is the ladder index of $\phi(x\,;z)$.
\end{proposition}

\begin{proof}
  Let $A\subseteq\Aa$ be finite.
  We prove that $A\subseteq\psi(\U\,;b_0,\dots,b_n)\subseteq\Aa$ for some $b_0,\dots,b_n$.
  To obtain the $b_k$, we construct a ladder inductively.
  Pick any $b_0$ be such that $A\subseteq\phi(\U\,;b_0)$, which we can do by approximability.
  Now, suppose that $a_1,\dots,a_k$ and $b_0,\dots,b_k$ have been defined.
  If 

  \ceq{\hfill A}{\subseteq}{\bigcap_{i=0}^k\phi(\U\,;b_i)\medrel{\subseteq}\Aa}

  we set $b_{k+1}=\dots=b_n=b_k$ and stop as we already have the required parameters.
  Otherwise pick any element

  \ceq{\hfill a_{k+1}}{\in}{\bigcap_{i=0}^k\phi(\U\,;b_i)\sm\Aa}

  and some $b_{k+1}$ such that

  \ceq{\hfill A}{\subseteq}{\phi(\U\,;b_{k+1})\medrel{\subseteq}\U\sm\{a_1,\dots,a_{k+1}\}}.

  Such a parameter $b_{k+1}$ exists because $\Aa$ is approximable (apply Definition~\ref{def_approx} with $A\cup\{a_1,\dots,a_{k+1}\}$ for $A$).
  Note that at stage $n$ we have constructed a ladder of length $n$ for $\neg\phi(x\,;z)$.
  In fact for all $h\in(n]$ and $k\in[n]$ we have
  
  \ceq{\hfill\neg\phi(a_h\,;b_k)}
  {\IFF}
  {h\le k.}
  
  As $\phi(x\,;z)$ has ladder index $n+1$, by Fact~\ref{fact_stability_neg} it is not possible to further prolong this chain, hence

  \ceq{\hfill A}{\subseteq}{\bigcap_{i=0}^n\phi(\U\,;b_i)\medrel{\subseteq}\Aa}

  as required.
\end{proof}

We prove that the formula $\psi(x\,;z_0,\dots,z_n)$ in the lemma above is itself stable, though with a larger ladder index.
  
\begin{lemma}
  Let $\psi_i(x\,;z)$, where $i=1,\dots,m$, be formulas with ladder index $n_i$. Let
  
  \ceq{\hfill \phi(x\,;z)}{=}{\bigwedge_{i=1}^m\phi_i(x\,;z)}
  
  Then $\phi(x\,;z)$ has ladder index $<R(n_1,\dots, n_m)$, the Ramsey number for $m$-colorings.
\end{lemma}
  
\begin{proof} 
  Suppose for a contradiction that there is a ladder $a_1,\dots,a_n$ and $b_0,\dots,b_n$ of length $n=R(n_1,\dots, n_m)$.
  Let $C_i$ contains the pairs $\{h,k\}$ such that $0\le k<h\le n$ and $\neg\phi_i(a_h\,;b_k)$.
  Then from \ssf{ld} we obtain 
  
  \ceq{\hfill\bigcup_{i=1}^mC_i}{=}{\binom{[n]}{2}}
  
  By the definition of $n$, for some $i\in[m]$, there is a set $H$ of cardinality $n_i$ such that $H^{(2)}\subseteq C_i$.
  Assume $i{=}1$ for definiteness.

  Write $a'_1,\dots,a'_{n_1}$ and $b'_0,\dots,b'_{n_1}$ for the tuples obtained by restricting $a_1,\dots,a_n$ and $b_0,\dots,b_n$ to the indexes in $H$.
  These tuples witness that $\phi_1(x\,;z)$ has ladder index at least $n_1+1$, which contradicts the assumption of the lemma.
\end{proof}

To apply the proposition to the formula $\psi(x\,;z_0,\dots,z_n)$ in the lemma above, let $z=z_0,\dots,z_n$ and let $\phi_i(x\,;z)=\phi(x\,;z_i)$.
That is, $\phi_i(x\,;z)$ defines a relation between $\U$ and $\V^{n+1}$ that depends only on the $i$-th coordinate.
The ladder index of every $\phi_i(x\,;z)$ is the same as $\phi(x\,;z_0)$ as a relation between $\U$ and $\V$.


\begin{lemma}
  If $\Aa$ is approximated from below by a stable formula $\phi(x\,;z)$ then for some $b_0,\dots,b_n\in\V$ 
  \\[-1ex]
  \ceq{\hfill \Aa}{=}{\bigcup^n_{i=0}\phi(\U\,;b_i)}
  
  where $n+1$ is the ladder index of $\phi(x\,;z)$. 
\end{lemma}
  
\begin{proof}
  Let $A\subseteq\U$ be finite.
  To obtain the $b_k$, we construct a ladder by stages.
  Pick any $b_0$ be such that $\phi(\U\,;b_0)\subseteq\Aa$.
  Now, suppose that $a_1,\dots,a_k$ and $b_0,\dots,b_k$ have been defined.
  If 
  \\[-1ex]
  \ceq{\hfill\Aa}{\subseteq}{\bigcup_{i=1}^k\phi(\U\,;b_i)}

  we set $b_{k+1}=\dots=b_n=b_k$ and stop.
  As the construction guarantees the converse inclusion, we have the required parameters.
  Otherwise pick any element

  \ceq{\hfill a_{k+1}}{\in}{\Aa\sm\bigcup_{i=1}^k\phi(\U\,;b_i)}

  and some $b_{k+1}$ such that

  \ceq{\hfill \{a_1,\dots,a_{k+1}\}}{\subseteq}{\phi(\U\,;b_{k+1})\medrel{\subseteq}\Aa}.

  Such a parameter $b_{k+1}$ exists because $\Aa$ is approximable from below.
  Note that at stage $n$ we have constructed a ladder of length $n$ for $\phi(x\,;z)$.
  In fact for all $h\in(n]$ and $k\in[n]$ we have
  
  \ceq{\hfill\neg\phi(a_h\,;b_k)}
  {\IFF}
  {h\le k.}
  
  As $\phi(x\,;z)$ has ladder index $n+1$, it is not possible to prolonge this chain, hence

  \ceq{\hfill\Aa}{=}{\bigcup_{i=0}^n\phi(\U\,;b_i)}

  as required.
\end{proof}
  
The main theorem of this section is an immediate corollary is the three lemmas above.

\begin{theorem}\label{thm_stable_definability}
  Let $\phi(x\,;z)$ be stable.
  Then there is a formula of the form

  \ceq{\hfill\psi(x\,;\bar z)}{=}{\bigvee_{j=0}^m\bigwedge_{i=0}^n\phi(z\,;z_{i,j})}
  
  such that every set $\Aa\subseteq\U$ approximable by  $\phi(x\,;z)$ is definable by $\psi(x\,;\bar b)$ for some $n{\times}m$-tuple of parameters $\bar b\in\V^{n\times m}$.\QED
\end{theorem}

We conclude this section with the sketch of an argument that shows that stable formulas have finite 2-rank, that is, \ssf{1}$\IMP$\ssf{2} of Theorem~\ref{thm_hodges}.

% Assume the 2-rank is infinite and reason for a contradiction.
% Assume for the moment that the rank is $\omega$, in the sense that there is an infinite branching tree $\<a_r:r\in2^{<\omega}\>$. 
% Let $\<b_s:s\in2^{\omega}\>$ be witnesses of \ssf{2r}.
% Let $Q\subseteq\{b_s:s\in2^{\omega}\}$ the set of sequences that are eventually $0$.
% Note that in the proof of Theorem~\ref{thm_stable_definability} we can require that $\bar b\in Q^{n\times m}$.
% But there are 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Stable Erd\H{o}s-Hajnal}

\end{document}
